\documentclass[12pt]{article}

\usepackage{amsmath}
\usepackage{geometry}
\geometry{margin=1in}

\setlength{\parskip}{0.5em}

\begin{document}

\title{Grade 11 Physics -- Kinematics Practice 2}
\author{}
\date{}
\maketitle

\vspace{1em}

\textbf{Instructions:}

\begin{enumerate}
\item Describe the motion before writing formulas.
\item Show all steps clearly.
\item Include units in every step and in the final answer.
\end{enumerate}

\noindent\rule{\textwidth}{0.4pt}

\newpage

\section*{Question 1: Reaction Time and Braking}

A car is traveling at a constant speed of $27.0 \, \text{m/s}$. The driver sees a stopped truck $95.0 \, \text{m}$ ahead.

The driver's reaction time is $1.20 \, \text{s}$. After this delay, the car brakes with a constant acceleration of $-6.00 \, \text{m/s}^2$.

\begin{itemize}
\item[(a)] Calculate the distance traveled during the reaction time.
\vspace{6cm}

\item[(b)] How far is the car from the truck when braking begins?
\vspace{6cm}

\item[(c)] Will the car collide with the truck? Justify your answer with calculations.
\vspace{8cm}

\end{itemize}



\newpage
\section*{Question 2: Two Cars Moving Toward Each Other}

Two cars are initially $240 \, \text{m}$ apart and move toward each other.

Car A has an initial velocity of $12.0 \, \text{m/s}$ and accelerates at $2.00 \, \text{m/s}^2$.

Car B has a constant velocity of $18.0 \, \text{m/s}$.

\begin{itemize}
\item[(a)] How long does it take for the cars to meet?
\vspace{8cm}

\item[(b)] How far from Car A's starting position do they meet?
\vspace{8cm}

\end{itemize}

\newpage

\section*{Question 3: Police Pursuit with Delay}

A car travels at a constant speed of $22.0 \, \text{m/s}$ past a stationary police car.

The police car waits $3.00 \, \text{s}$ before accelerating at $2.50 \, \text{m/s}^2$.

\begin{itemize}
\item[(a)] How far ahead is the car when the police begins accelerating?
\vspace{7cm}

\item[(b)] How long does it take for the police car to catch up (after it starts accelerating)?
\vspace{9cm}

\item[(c)] What is the police car's velocity when it catches the car?
\vspace{6cm}

\end{itemize}

\newpage

\section*{Question 4: Object Thrown Downward}

A ball is thrown downward from the top of a $60.0 \, \text{m}$ building with an initial velocity of $5.00 \, \text{m/s}$ downward.

Assume acceleration due to gravity is $9.8 \, \text{m/s}^2$ downward.

\begin{itemize}
\item[(a)] How long does it take for the ball to reach the ground?
\vspace{8cm}

\item[(b)] What is the velocity of the ball when it hits the ground?
\vspace{8cm}

\end{itemize}

\newpage

\section*{Question 5: Ball Thrown Upward}

A ball is thrown straight upward from the ground with an initial velocity of $18.0 \, \text{m/s}$.

Assume acceleration due to gravity is $9.8 \, \text{m/s}^2$ downward.

\begin{itemize}
\item[(a)] Calculate the maximum height reached by the ball.
\vspace{8cm}

\item[(b)] How many times does the ball pass the height of $12.0 \, \text{m}$?
\vspace{6cm}

\item[(c)] Calculate the velocity at $12.0 \, \text{m}$ for each time it passes this height. Include direction.
\vspace{10cm}

\end{itemize}

\end{document}